Higher Education Synthesized

### 2025-02-08 report/msds_data_mining_project_report.tex
finish first pass of project report, before starting technical work

\section{Introduction}

    American higher education saw growing enrollments throughout the latter half of the 1900s and moving into the 200s. This resulted in increases in the number of higher education institutions (HEIs) in the country, including public and private colleges and universities. These HEIs span from major research institutions down to small liberal arts schools. This has also resulted in increases in the number of faculty and staff to support growing enrollments. The general trend in higher education was growth.

@@

    \cite{DBLP:journals/corr/abs-1907-00503}, by the same authors as the previous TGAN paper, provides more consideration for nuances and potential issues in the data along with adding a deep conditional GAN (DCGAN) as part of their new conditional tabular GAN (CTGAN) model. DCGAN generally is a GAN architecture that uses convolutional neural networks (CNNs) in both the generator and discriminator. One example of extra data considerations is that CTGAN handles severe imbalance in categorical features. We will not worry about the detail of this model either, but it is potentially one that we will use with the modeling library we introduce next

@@

\section{Proposed Work}

    For me, at this stage, this is the section that I am looking to narrow down a bit (and I am open to suggestions from readers). I have a couple of ideas planned for the work in this project, and I am looking to lean into my background with the architecting of data systems to build out a wider pipeline.

@@

\section{Evaluation}

    Tests for synthetic data
    data mining algorithms

Evaluation metrics
    - Effectiveness
    - Efficiency
Experimental setup
Methods to compare

\section{Discussion}
Project timeline
    - When to finish
    - Current status
    - Determine if you need to do less or can do more in the given time
Potential challenges
Alternative approaches and backup plan

\section{Conclusion}
Conclusion
Project summary
Key findings (when finished)
Future work (when finished)

@@

\section{Technical Details}

    Need some details plz

>>>>

\section{Introduction}
    American higher education saw growing enrollments throughout the latter half of the 1900s and moving into the 200s. This resulted in increases in the number of higher education institutions (HEIs) in the country, including public and private colleges and universities. These HEIs span from major research institutions down to small liberal arts schools. This has also resulted in increases in the number of faculty and staff to support growing enrollments. The general trend in higher education was growth.

@@

    \cite{DBLP:journals/corr/abs-1907-00503}, by the same authors as the previous TGAN paper, provides more consideration for nuances and potential issues in the data along with adding a deep conditional GAN (DCGAN) as part of their new conditional tabular GAN (CTGAN) model. DCGAN generally is a GAN architecture that uses convolutional neural networks (CNNs) in both the generator and discriminator. One example of extra data considerations is that CTGAN handles severe imbalance in categorical features. We will not worry about the detail of this model either, but it is potentially one that we will use with the modeling library we introduce next.

@@

\section{Proposed Work}
    For me, at this stage, this is the section that I am looking to narrow down a bit (and I am open to suggestions from readers). I have a couple of ideas planned for the work in this project, and I am looking to lean into my background with the architecting of data systems to build out a wider pipeline.

@@

    I have spent most of the time here focusing on the technical side of the work. I need to dig into the IPEDS data before feeling more comfortable with the questions in terms of meaning that we will explore, but I expect those to center around identifying patterns in clusters of institutions based on institution attributes, enrollment trends, and graduation trends, possibly even dropping either enrollment or graduation data in favor of simplifying the data scope for right now.

\section{Evaluation}
    Until we decide to drop synthetic data, if we do make that decision, the first evaluations we will explore are how similar the synthetic data is to the real data. We will explore the evaluation metrics in the Synthetic Data Vault Project. A couple of candidates that stand out are per-attribute comparisons and machine learning score. Per-attribute comparisons check the same feature in the real and synthetic datasets, calculate metrics such as mean or standard deviation, and determine the quality of the synthetic data based on how close these measures are across the datasets.

    Machine learning score trains classifiers on the real and synthetic datasets and compares the performance of those models. For example, if we use both classifiers to check something like accuracy or F1 score on a held-out test set, we can evaluate how similar the synthetic data is to the real data based on comparing those scores.

    We can also look at more advanced evaluation metrics. One that has stood out to me predicting each feature in a dataset with all of the other features. If we find the predictive power in the real and synthetic datasets are similar, then, like above, that can give us a sense of how similar the datasets themselves are.

    There are any number of other evaluative metrics for synthetic data. We will explore what is available with the project we use for synthetic data and will clarify here as we learn more during experimentation.

    Moving on from synthetic data, until we explore the IPEDS data further, I am not sure exactly what we will end up modeling or testing as output. But there are some general contenders that we could likely look at.

    For supervised modeling outputs, we can look at standard metrics such as accuracy, precision, F1 score, and more for classification problems. We can use a metric like mean squared error (MSE) to measure predictive model output. These evaluations give us firm numbers we can use to determine how well our modeling is doing.

    But, I am actually more intrigued by seeing what we can find with unsupervised learning. Admittedly, the outputs for unsupervised learning can be more difficult to work with, especially at this early stage. Clustering is the first unsupervised modeling that comes to mind. I often think of clustering as a much more advanced version of EDA. We are not really sure what patterns exist in the data, so we supercharge standard EDA and see what patterns we uncover when leveraging machine learning. But we do not have a clean evaluation metric with clustering. We will need to interactively explore the clustering outputs, at least at first. At some point, we can determine a set number of clusters and pivot the clustering into a classification approach, but the more interesting piece here for me is the clustering itself.

    I am on the fence about leaning more into the synthetic data approach in which we would spend a lot more time evaluating the quality of the synthetic data. With synthetic data, I still would like to test out one of the analytical pieces at the end of the pipeline, but maybe we stick to something simpler like a clearer classification problem. If we instead drop synthetic data, then we have more room to tackle something like unsupervised learning and the complexity inherent in it.

    While I feel like there are loads of evaluative methods to explore, I am also interested in looking at some of the algorithms from the second data mining course. I have leveraged things like similarity scores in other projects, but I have not experimented with, for instance, FP growth, and it would be interesting to possibly pick a brand new algorithm to plop at the end of the pipeline and figure out what it means to evaluate that algorithm and what it reveals to us.

    So, at this stage, this section is more of a brainstorming of some of the evaluative approaches. This will get refined down in the next iteration after exploring the IPEDS data a bit and determining if we will keep or drop the synthetic data component of the pipeline.

\section{Discussion}
    I tend to drag out projects, taking the opportunity to add increased complexity and explore new items I am not as familiar with. For this project, part of me wants to focus on sticking a bit more in-bounds in terms of my comfort level with data pipelines, only adding in a little extra complexity and seeing if it is feasible to build the entire pipeline out in a shorter amount of time. While spending forever on projects is fun and educational, I do risk getting too comfortable with taking my time. I could benefit from descoping and focusing on getting smaller chunks working earlier and then deciding if it is worth expanding out from there.

    In that vein, I am hoping that the refinement in going from the first to second draft of this report, spending some time to explore the technical and data sides of the project in between, will help reveal scope from this draft that can be cut. It would be ideal if I could use a three or four week timeline and adjust the project to fit within that, but it does mean that what I have laid out above is probably too ambitious for that timelines. But, after writing this all out, the scope does feel pretty big at the moment.

    Currently I have identified some IPEDS data, but I have not written any code or set up the database. My approach will be to get the data through to at least the initial EDA and preprocessing before starting to evaluate the scope after that. I feel comfortable with the earlier and more technical steps in terms of time, so my question marks sit with the longer EDA, the modeling, and, if included, the synthetic data portion.

    The main challenge I see is that the data ends up being much more complicated to work with than expected, so I will need to shift the scope down even further to prioritize working with difficult data. That would not be a bad way for the project to pivot, but it is not what I am planning on at this stage. If the data turns out to be friendly enough to work with, then I think the next big challenge will be figuring out the right analysis or modeling in terms of the meaning of the data. The ideas above are all interesting from a technical standpoint, but the meaning of what we dig into needs to drive why we are using those approaches.

    The general backup plan is to limit each stage of the pipeline to maybe one or two pieces. This may mean we do some simple EDA along with a correlation matrix, we look at a clearer regression or classification problem, or we leverage synthetic data but trim the input dataset to only a few features. All of these are tactics for trimming the scope, likely the biggest risk for the project at the moment.

\section{Conclusion}
    Not much to conclude at the moment. We will fill this in as we start doing the data work of the project. What I expect to see here is that we trimmed down the IPEDS data, scoped the use case for the project to be a bit narrower, decided to explore synthetic data, and picked a simpler overall output. For a surprise finish, we may end up deciding to pull some data out and run it through a separate business intelligence tool like Tableau with the idea that end users of the production pipeline may want to leverage tools such as that for their purposes.

    I expect the next steps to end up being that we incorporate more IPEDS data to answer a broader range of questions. I do not expect radical changes in the pipeline architecture in next iterations of the project, but I do expect additions and expansions.

@@

\section{Technical Details}
    Need some details plz. Leaving this as a placeholder to pull the technical details out of the narrative above but still keep it available for readers.
